Anthropogenic plant communities of the Iberian Atlantic ecoregion

### 2024-04-11 doc/manuscript.Rmd
revised methods

---
title: "Anthropogenic plant communities of the Iberian Atlantic ecoregion"
output: 
  bookdown::word_document2:
    reference_docx: template.docx
    number_sections: false
bibliography: refs.bib
csl: journal-of-vegetation-science.csl
link-citations: yes
always_allow_html: true
---

@@

@@

We studied the anthropogenic plant communities of the Iberian Atlantic ecoregion, i.e. the territories with a temperate climate in the north-western Iberian Peninsula. Our study ecoregion broadly corresponds with the Cantabrian Mixed Forests ecoregion *sensu* Olson *et al.* [-@RN4667], with the Iberian part of the European Atlantic province *sensu* Rivas-Martínez *et al.* [-@RN4699] and with the Iberian section of the Atlantic biogeographical region of the European Environmental Agency (https://www.eea.europa.eu/data-and-maps/figures/biogeographical-regions-in-europe-2). To define in detail the study ecoregion we did a fine-scale revision of its geographic borders at high spatial resolution using the most recent biogeographical map of the Iberian Atlantic territories [@RN5372]. The revised Iberian Atlantic ecoregion encompasses 79,891 km^2^ of Portugal, Spain and France. Its climate is temperate with average monthly temperatures ranging from 6 to 20°C, and with total annual precipitation between 1,100 and 2,600 mm, with some areas having warm-dry summers but without the distinctive 2-month drought of the Mediterranean climate [@RN5373]. Nonetheless, the Iberian Atlantic ecoregion is located at the transition between the temperate and the Mediterranean climates in southern Europe.

@@

To define our target vegetation, we followed the definitions of anthropogenic vegetation in the recent classification of the vegetation of Europe by Mucina *et al.* [-@RN4698]. In our study area, this potentially includes the vegetation classes *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Chenopodietea*, *Sisymbrietea*, *Artemisietea vulgaris*, *Epilobietea angustifolii* and *Bidentetea*. For a complete synthesis, we added the class *Cymbalario-Parietarietea diffusae* included by Mucina *et al.* [-@RN4698] in the vegetation of rock crevices and screes, since this class encompasses the vegetation of human-made walls in our ecoregion. For the nomenclature of sintaxa we follow Mucina *et al.* [-@RN4698] for alliances and higher ranks, and Rivas-Martínez *et al.* [-@RN4697] for associations.

## Vegetation data

We retrieved anthropogenic vegetation plots from SIVIM, the Iberian and Macaronesian Vegetation Information System [@RN5374]. To identify plots as anthropogenic we followed two criteria: (1) We used the expert system created by Chytrý *et al.* [-@RN5061] to classify vegetation plots into EUNIS pan-European habitat types. This expert system uses the floristic composition of plots to assign them to habitat types depending on the presence and abundance of characteristic species. We used the habitats assigned by the expert system to retrieve plots that had been assigned to any habitat related to the vegetation classes we had previously defined: all habitats in the level 1 code V (vegetated man-made habitats), plus level 3 codes R55 (lowland moist or wet tall-herb and fern fringe) and R57 (herbaceous forest clearing vegetation) (R55 and R57 include communities that are classified as "grasslands" by EUNIS but as "anthropogenic" by by Mucina *et al.* [-@RN4698]). (2) We also retrieved vegetation plots from SIVIM if they had been classified by the original authors of the relevé into any sintaxa belonging to our target anthropogenic vegetation classes (**Supplementary 1**). This produced an initial pool of 3,160 vegetation plots from SIVIM, to which we added 89 vegetation plots of urban or peri-urban plant communities sampled by us or extracted from local literature not in SIVIM [@RN5377; @RN5378]. After initial exploration of this dataset using Two-Way Indicator Species Analysis (TWINSPAN) [@RN5379], we identified 652 outlier plots which either did not belong to the target anthropogenic classes or to the geographical area of the ecoregion. The majority of these plots corresponded to coastal plant communities and had been misclassified by the expert system. We removed these, leaving a dataset of 2,508 vegetation  plots for data analyses: 2,419 plots from SIVIM, originally recorded in 89 publications (**Supplementary 2**); plus 89 plots added by us.

## Statistical analyses

We performed all data analyses with R version 4.3.1 [@RN5387], using the R package 'tidyverse' [@RN4662] for data processing and visualization. The original datasets, as well as R code for analysis and creation of the manuscript can be accessed at the GitHub repository https://github.com/efernandezpascual/manmade. We homogenized all taxon names in the analyses and the article using Euro+Med [@RN5044] or Plants of the World Online for taxa not in Euro+Med [@RN5376].

[Definition of target alliances and training dataset.]{.underline} As a previous step before classification, we prepared a revised checklist of potential anthropogenic alliances that could be present in our study ecoregion, based on regional sintaxonomical checklists [@RN5363; @RN5008; @RN5365] and recent revisions at the European level [@RN4698; @RN5371]. Then, we filtered our dataset to extract a subset with the vegetation plots which had been assigned by the authors to any of these alliances. We used TWINSPAN to conduct a preliminary numerical classification of this subset, attempting to match the plots into the *a priori* defined checklist of alliances. We used 3 pseudospecies cut levels (0, 15, 25) with a minimum group size of 10 plots and Sørensen's average dissimilarity, and the R package 'twinspanR' [@RN5380] to do the TWINSPANs. To start, we did a TWINSPAN classification into 15 clusters, followed by separate TWINSPANS within each cluster. This preliminary TWINSPAN allowed us to (1) validate which alliances were present in the study ecoregion; and (2) prepare a training dataset for semi-supervised classification (see below).

[Semi-supervised classification.]{.underline} Next, we applied a semi-supervised classification to the whole dataset. Semi-supervised classification uses a training subset of *a priori* classified vegetation plots to classify a secondary subset of unclassified plots [@RN5383]. Our training subset included all the plots that (1) had been classified by the authors of the relevé into any of the alliances that we had validated during the previous TWINSPANs, and (2) had also been matched by us into these alliances during our TWINSPANs. The secondary subset included all the plots which did not meet these two criteria. Our goal was to refine and not to simply extend the classification, so we applied the semi-supervised classification to the whole dataset, i.e. we allowed plots from the training subset to be re-assigned to other groups during the classification. In addition, semi-supervised classification can create new groups to place data points that do not match the already existing *a priori* groups. Furthermore, we applied a noise clustering fuzzy algorithm, which allows the plots to be classified either into the *a priori* groups or into a 'noise' group which includes outliers and transitional plots [@RN5382]. We set the fuzziness coefficient to a low value (m = 1) to accommodate a high number of transitional plots; and we set the distance to the noise class to d = 1. We performed this analysis with the R package 'vegclust' [@RN5383]. We used pairwise PERMANOVA (with 100,000 iterations, Euclidean distances and Holm's p-value correction) fitted with the R package 'RVAideMemoire' [@RN5381] to test the significance of the final vegetation groups; along with Detrended Correspondence Analysis (DCA) as implemented in the R package 'vegan' [@RN5386] to visualize the relationships between the groups.

[Characteristic species and EUNIS habitat regionalization.]{.underline} We calculated sets of characteristic species [@RN5061] for each of the final vegetation groups. We defined dominant species as species with more than 25% cover in at least 5% of the vegetation plots of the group; constant species as species with a frequency higher than 50% in the group; and diagnostic species as species with an 'IndVal' value with a p-value lower than 0.05, as calculated with the R package 'labdsv' [@RN2311] using 1.000.000 iterations. Additionally, we assigned to each group a regionalized level 4 EUNIS habitat code, using pre-existing codes when they existed (https://eunis.eea.europa.eu/habitats-code-browser-revised.jsp), and creating new codes when not.

[Community traits and ecological requirements.]{.underline} To describe the resulting vegetation groups, we compiled a dataset of species traits and ecological preferences for the species pool of the classified vegetation plots. For each species, we extracted from *Flora iberica* [-@RN3243] its maximum height, its median month of flowering, the length of the flowering period, and whether it was a neophyte or not. From FloraVeg.EU [-@RN5385] we extracted the species’ life forms and kept 'therophyte' and 'geophyte' for further analysis since these two categories had the largest contribution to variance as per an exploratory Principal Component Analysis (PCA) performed using the R package 'FactoMineR' [@RN3166]. We also collected the ecological indicator values [@RN5384] and the disturbance indicator values [@RN5101] of each species. In the case of the ecological and disturbance indicators, we assigned values to species lacking them by (1) calculating weighted average values of the species with values for each plot (weighting by species cover), and (2) assigning to the missing species the weighted average values of the plots where they were present (weighting by species cover). Then, we calculated the plot-level community-weighted mean (weighting by species cover) for each trait and ecological preference, and explored the main axes of variation using PCA, separately for species traits and ecological preferences.

@@

Our revision of the literature produced a provisional list of 38 anthropogenic vegetation alliances that could be present in the Iberian Atlantic ecoregion (**Supplementary 3**). In our study dataset (n = 2,508 vegetation plots), there were 2,201 plots that had been assigned by the original authors to one of these alliances. We used these 2,201 plots for a preliminary re-classification based on TWINSPAN, attempting to re-match the plots into the *a priori* defined checklist of 38 alliances. This procedure allowed us to validate 28 alliances as present in the ecoregion. A total of 1,725 plots had been classified by both the original authors and our TWINSPAN into these alliances, and these became our training subset for the semi-supervised classification. Attempts by the semi-supervised classification to create new groups resulted in new groups with no ecological significance, and thus we kept the original 28 alliances as final vegetation groups (**Table 1**). Most of the alliances had a wide distribution in the ecoregion according to our dataset (**Figure 1**), although five of them (*Caucalidion lappulae*, *Linario polygalifoliae-Vulpion alopecuri*, *Convolvulo arvensis-Agropyrion repentis*, *Senecionion fluviatilis*, *Paspalo-Agrostion semiverticillati*) had only isolated occurrences in a few points towards the southern limit of the Iberian Atlantic territories. The regional literature included in our dataset recognized 72 associations (**Supplementary 4**) within these alliances. The semi-supervised classification with noise clustering resulted in the final classification of 2,081 plots into 28 alliances and 427 plots left out in the noise group.

@@

@@

16. *Convolvulo arvensis-Agropyrion repentis* Semiruderal grasslands. Occupies wetter and shadier situations than the other alliances in the class, and has a higher proportion of geophytes and grasses (**Fig. 4**). Isolated occurrence in the south of the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Poa compressa*, *Elytrigia repens*, *Potentilla reptans*, *Valerianella locusta*.

17. *Carduo carpetani-Cirsion odontolepidis* Thistle ruderal vegetation of warm sites. Occupies warmer situations than the other thistle alliance in the class (**Fig. 4**). Mostly distributed in the Cantabrian Mountains (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Onopordum acanthium*.

18. *Cirsion richterano-chodati* Thistle ruderal vegetation of cold sites. Occupies colder situations than the other thistle alliance in the class (**Fig. 4**). Mostly distributed in the Cantabrian Mountains (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Cirsium eriophorum*, *Carduus carpetanus*, *Carduus nutans* aggr.

19. *Dauco-Melilotion* Biennial ruderal vegetation. Occupies wetter situations than the two thistle alliances in the class, and has a higher proportion of therophytes (**Fig. 4**). Sparse occurrence throughout the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Helminthotheca echioides*, *Daucus carota*, *Melilotus albus*, *Foeniculum vulgare*.

@@

20. *Geo urbani-Alliarion officinalis* Low-herb short-lived semiruderal and fringe vegetation. Occupies shadier and less disturbed situations than the other alliances in the class, and has a higher proportion of therophytes and shorter plants (**Fig. 4**). Wide distribution in the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Urtica dioica*.

21. *Arction lappae* Low-herb short-lived ruderal vegetation. Occupies sunnier and more frequently-disturbed situations than the other alliances in the class, and has a higher proportion of therophytes and shorter plants (**Fig. 4**). Distribution concentrated in the Cantabrian Mountains (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Oxybasis rubra*, *Senecio duriaei*.

22. *Balloto-Conion maculati* Tall-herb perennial ruderal vegetation. Occupies warmer situations than the other alliances in the class (**Fig. 4**). Distribution concentrated in the Cantabrian Mountains and the east of the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Sambucus ebulus*.

23. *Aegopodion podagrariae* Tall-herb perennial vegetation of forest margins and clearings. Occupies shadier and less disturbed situations than the other alliances in the class (**Fig. 4**). Sparse occurrence throughout the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Anthriscus sylvestris*, *Heracleum sphondylium*, *Galium aparine*.

24. *Epilobion angustifolii* Tall-herb perennial vegetation of forest margins and clearings in acidic soils. Occupies colder, nutrient-poorer, more acidic and less disturbed situations than the other alliances in the class (**Fig. 4**). Distributed towards the west of the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Epilobium angustifolium*, *Digitalis purpurea*, *Luzula lactea*.

25. *Cynancho-Convolvulion sepium* Tall-herb vegetation of nutrient-rich riparian habitats. Occupies wetter situations and less disturbed situations than the other alliances in the class, and has a later flowering season (**Fig. 4**). Dispersed throughout the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Eupatorium cannabinum*, *Angelica sylvestris*, *Picris hieracioides*.

26. *Senecionion fluviatilis* Tall-herb vegetation of nutrient-rich river banks and ditches. Occupies warmer and wetter situations than the other alliances in the class, flowers later, has taller plants and is highly dominated by neophytes and geophytes (**Fig. 4**). Isolated occurrence in the western coast of the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Arundo donax*, *Silene latifolia*.

27. *Bidention tripartitae* Summer-annual pioneer vegetation of temporarily-flooded sites. Occupies nutrient-richer sites than the other alliance in the class and has a higher proportion of therophytes (**Fig. 4**). Dispersed throughout the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Persicaria hydropiper*, *Bidens frondosus*, *Lythrum salicaria*.

28. *Paspalo-Agrostion semiverticillati* Summer-annual pioneer vegetation of temporarily-flooded and warm sites. Occupies warmer and wetter situations than the other alliance in the class, has a higher proportion of neophytes and has a longer flowering season (**Fig. 4**). Isolated occurrences in the south of the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Paspalum distichum*, *Cyperus longus*, *Schoenoplectus lacustris*.

@@

Compare proportion of neophytes to other areas [@RN5429; @@RN5435]

@@

@@

The original datasets, as well as R code for analysis and creation of the manuscript can be accessed at the GitHub repository https://github.com/efernandezpascual/manmade. Upon publication, a version of record of the repository will be deposited in Zenodo.

@@

```{r fig1, echo = FALSE, fig.pos = "H", fig.cap = "Figure 1. Distribution of the anthropogenic alliances in the Iberian Atlantic ecoregion. Each dot represents a relevé. Dots are colored by vegetation classes."}
knitr::include_graphics("../results/figures/F1 - maps.png")
```

```{r fig2, echo = FALSE, fig.pos = "H", fig.cap = "Figure 2. Ordination of the floristic composition of the anthropogenic plant communities of the Iberian Atlantic ecoregion. The ordination is the results of a Detrended Correspondence Analysis grouped by vegetation classes (A, left) and alliances (B, right)."}
knitr::include_graphics("../results/figures/F2 - pca.png")
```

```{r fig3, echo = FALSE, fig.pos = "H", fig.cap = "Figure 3. Ordination of the community-level species traits and ecological preferences. The ordinations are the results of a Principal Component Analysis calculated for the community-weighted means for species traits (A, left) and ecological and disturbance indicator values (B, right)."}
knitr::include_graphics("../results/figures/F3 - origins.png")
```

```{r fig4, echo = FALSE, fig.pos = "H", fig.cap = "Figure 4. Community-level species traits and ecological preferences by alliance. Boxplots calculated for the community-weighted means for species traits and ecological and disturbance indicator values."}
knitr::include_graphics("../results/figures/F4 - lifeforms.png")
```

```{r fig5, echo = FALSE, fig.pos = "H", fig.cap = "Figure 4. Community-level species traits and ecological preferences by alliance. Boxplots calculated for the community-weighted means for species traits and ecological and disturbance indicator values."}
knitr::include_graphics("../results/figures/F5 - height-flowering.png")
```

```{r fig6, echo = FALSE, fig.pos = "H", fig.cap = "Figure 4. Community-level species traits and ecological preferences by alliance. Boxplots calculated for the community-weighted means for species traits and ecological and disturbance indicator values."}
knitr::include_graphics("../results/figures/F6 - indicators.png")
```

```{r fig7, echo = FALSE, fig.pos = "H", fig.cap = "Figure 4. Community-level species traits and ecological preferences by alliance. Boxplots calculated for the community-weighted means for species traits and ecological and disturbance indicator values."}
knitr::include_graphics("../results/figures/F7 - pca.png")
```

>>>>

---
title: "Anthropogenic plant communities of the Iberian Atlantic ecoregion"
output: 
  bookdown::github_document2:
    reference_docx: template.docx
    number_sections: false
bibliography: refs.bib
csl: journal-of-vegetation-science.csl
link-citations: yes
always_allow_html: true
---

@@

We performed all data management and analysis with R version 4.3.1 [@RN5387], using the R package *tidyverse* [@RN4662] for data processing and visualization. We homogenized all taxon names in the analysis and the article using Euro+Med [@RN5044] or Plants of the World Online for taxa not in Euro+Med [@RN5376].  For the nomenclature of sintaxa we follow Mucina *et al.* [-@RN4698] for alliances and higher ranks, and Rivas-Martínez *et al.* [-@RN4697] for associations. The original datasets, as well as R code for analysis and creation of the manuscript can be accessed at the GitHub repository <https://github.com/efernandezpascual/manmade>.

@@

We studied the anthropogenic plant communities of the Iberian Atlantic ecoregion [@RN5372], i.e. the territories with a temperate climate in the north-western Iberian Peninsula. Our study ecoregion broadly corresponds with the Cantabrian Mixed Forests ecoregion *sensu* Olson *et al.* [-@RN4667], with the Iberian part of the European Atlantic province *sensu* Rivas-Martínez *et al.* [-@RN4699] and with the Iberian section of the Atlantic biogeographical region of the European Environmental Agency (<https://www.eea.europa.eu/data-and-maps/figures/biogeographical-regions-in-europe-2>).

@@

To circumscribe our study vegetation, we followed the definition of anthropogenic vegetation in the recent classification of the vegetation of Europe by Mucina *et al.* [-@RN4698]. In our study area, this potentially includes the vegetation classes *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Chenopodietea*, *Sisymbrietea*, *Bidentetea*, *Artemisietea vulgaris* and *Epilobietea angustifolii*. To these, and for a complete synthesis, we added to our study the class *Cymbalario-Parietarietea diffusae* included by Mucina *et al.* [-@RN4698] in the vegetation of rock crevices and screes, since this class encompasses the vegetation of human-made walls.

## Checklist of anthropogenic sintaxa

To assist in our classification, we prepared a checklist of anthropogenic sintaxa that could potentially be present in our study ecoregion (**Supplementary 1**), based on regional sintaxonomical checklists [@RN5363; @RN5008; @RN5365] and recent revisions at the European level [@RN4698; @RN5371]. The checklist inclused 38 anthropogenic vegetation alliances that could be present in the Iberian Atlantic ecoregion (**Supplementary 2**). A goal of our study was to validate the presence of these alliances in our ecoregion, and to assign to them the existing vegetation data.

## Vegetation data selection

We retrieved anthropogenic vegetation plots from SIVIM, the Iberian and Macaronesian Vegetation Information System [@RN5374]. To identify plots as anthropogenic, we selected those plots that had been classified by the original authors of the plot as belonging to any sintaxa in our sintaxonomical checklist. However, since not all plots in SIVIM have an assigned sintaxon, we also retrieved additional plots using the expert system created by Chytrý *et al.* [-@RN5061] to classify vegetation plots into EUNIS pan-European habitat types. This expert system uses the floristic composition of plots to assign them to habitat types depending on the presence and abundance of characteristic species. We retrieved plots that had been assigned to any habitat related to the vegetation classes we had previously defined: all habitats in the level 1 code V (vegetated man-made habitats), plus level 3 codes R55 (lowland moist or wet tall-herb and fern fringe) and R57 (herbaceous forest clearing vegetation). It must be noted that habitats R55 and R57 include communities that are classified as *grasslands and lands dominated by forbs, mosses or lichens * by EUNIS but as *anthropogenic vegetation* by by Mucina *et al.* [-@RN4698]. The extraction from SIVIM produced an initial pool of 3,160 vegetation plots, to which we added 89 vegetation plots of urban or peri-urban plant communities sampled by us or extracted from local literature not in SIVIM [@RN5377; @RN5378].

## Vegetation data cleaning

We performed an exploratory analysis of these 3,249 plots using Two-Way Indicator Species Analysis (TWINSPAN) [@RN5379]. We used the R package *twinspanR* [@RN5380] to do the TWINSPANs in this article, using 3 pseudospecies cut levels (0, 15, 25) with a minimum group size of 10 plots and Sørensen's average dissimilarity. During this exploratory analysis we identified 741 outlier plots which either did not belong to the target anthropogenic vegetation or to the geographical area of the ecoregion. The majority of these plots corresponded to coastal plant communities and had been misclassified by the expert system. We removed these, leaving a dataset of 2,508 vegetation plots for data analysis: 2,419 plots from SIVIM, originally recorded in 89 publications (**Supplementary 3**); plus 89 plots added by us.

## Training dataset for semi-supervised classification and validation of alliances

To create a training dataset for semi-supervised classification (see below), we started by keeping only those plots (n = 2,201) that had been assigned by the original authors of the plot as belonging to any of the 38 alliances in our sintaxonomical checklist (i.e., we removed the plots without sintaxa, that had been retrieved solely by the expert system). We further cleaned this subset by subjecting it to a preliminary numerical classification using TWINSPAN, in which we attempted to match the plots into the 38 alliances our sintaxonomical checklist. This step allowed us to identify those plots in which there was an agreement between (1) our TWINSPAN-based classification and (2) the classification based on the original-author sintaxa; these plots (n = 1,725) became our training dataset for semi-supervised classification. Furthermore, this step allowed us to validate which alliances from our sintaxonomical checklist were present in the study ecoregion (n = 28) and which were absent (n = 10).

## Semi-supervised classification

Next, we conducted a semi-supervised classification of the whole dataset (n = 2,508 plots) into the 28 validated anthropogenic alliances. Semi-supervised classification uses a training subset of *a priori* classified vegetation plots to classify a secondary subset of unclassified plots [@RN5383]. Since our goal was to refine the classification of the whole dataset, we allowed plots from the training subset to be re-assigned to other alliances during the classification. In addition, semi-supervised classification can create new groups to place data points that do not match the already existing *a priori* groups, but attempts to do so resulted in new groups with no ecological significance, and thus we kept the  28 alliances as final vegetation groups. Furthermore, we applied a noise clustering fuzzy algorithm, which allows the plots to be classified either into the *a priori* alliances or into a *noise* group which includes outliers and transitional plots [@RN5382]. We set the fuzziness coefficient to a low value (m = 1) to accommodate a high number of transitional plots; and we set the distance to the noise class to d = 1. We performed this analysis with the R package *vegclust* [@RN5383]. The semi-supervised classification with noise clustering resulted in the final classification of 2,081 plots into 28 alliances and 427 plots left out in the noise group. We used pairwise PERMANOVA (with 100,000 iterations, Euclidean distances and Holm's p-value correction) fitted with the R package *RVAideMemoire* [@RN5381] to test the significance of the final vegetation alliances; along with Principal Component Analysis (PCA) as implemented in the R package *FactoMineR* [@RN3166] to visualize the relationships between the alliances.

## Characteristic species and EUNIS habitat regionalization

We calculated sets of characteristic species [@RN5061] for each of the final 28 vegetation alliances. We defined dominant species as species with more than 25% cover in at least 5% of the vegetation plots of the group; constant species as species with a frequency higher than 50% in the group; and diagnostic species as species with an *IndVal* value with a p-value lower than 0.05, as calculated with the R package *labdsv* [@RN2311] using 1.000.000 iterations. Additionally, we assigned to each alliance a regionalized level 4 EUNIS habitat code, using pre-existing codes when they existed (<https://eunis.eea.europa.eu/habitats-code-browser-revised.jsp>), and creating new codes when not.

## Species origin as native, archaeophyte and neophyte

We classified the species as native, archaophytes and neophytes using the information in *Flora iberica* [-@RN3243] and catalogues of archaeophytes for Britain [@RN5490] and the Czech Republic [@RN5495]. We must stress that identifying archaeophytes is southern Europe is highly problematic and our classification must be taken as an indication of putative archaeophyte character for the purposes of vegetation description, rather than a definitive classification of the species.

## Species life form

From *FloraVeg.EU* [-@RN5385] we extracted the species’ life forms and kept *therophyte* and *geophyte* for further analysis since these two categories had the largest contribution to variance as per an exploratory Principal Component Analysis (PCA) performed using the R package *FactoMineR* [@RN3166].

## Species and vegetation height and flowering phenology

From *Flora iberica* [-@RN3243] we extracted the maximum height, median month of flowering and length of the flowering period of each species in the dataset. We used these to calculate plot-level community-weighted means (weighting by species cover) for each trait, to characterize the vegetation height and flowering phenology of the anthropogenic plant communities.

## Species ecological indicator values

We collected, for each species in the dataset, its ecological indicator values of temperature, moisture, light, nutrients and soil reaction [@RN5384]; and its disturbance frequency and severity indicator values [@RN5101]. To those species lacking an indicator value in the accessed references, we assigned a value by reciprocal averaging, i.e. by (1) calculating weighted average values of the species with values for each plot (weighting by species cover), and (2) assigning to the missing species the weighted average values of the plots where they were present (weighting by species cover). Then, for each vegetation plot, we calculated the plot-level mean of each indicator value.

@@

The semi-supervised classification resulted in the classification of 2,081 vegetation plots into 28 anthropogenic alliances of the Iberian Atlantic ecoregion (**Table 1**, **Figure 1**). Most of the alliances had a wide distribution in the ecoregion according to our dataset, although five of them (*Caucalidion lappulae*, *Linario polygalifoliae-Vulpion alopecuri*, *Convolvulo arvensis-Agropyrion repentis*, *Senecionion fluviatilis*, *Paspalo-Agrostion semiverticillati*) had only isolated occurrences in a few points towards the southern limit of the Iberian Atlantic territories. The regional literature included in our dataset recognized 72 associations (**Supplementary 4**) within these alliances.

@@

16. *Bidention tripartitae* Summer-annual pioneer vegetation of temporarily-flooded sites. Occupies nutrient-richer sites than the other alliance in the class and has a higher proportion of therophytes (**Fig. 4**). Dispersed throughout the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Persicaria hydropiper*, *Bidens frondosus*, *Lythrum salicaria*.

17. *Paspalo-Agrostion semiverticillati* Summer-annual pioneer vegetation of temporarily-flooded and warm sites. Occupies warmer and wetter situations than the other alliance in the class, has a higher proportion of neophytes and has a longer flowering season (**Fig. 4**). Isolated occurrences in the south of the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Paspalum distichum*, *Cyperus longus*, *Schoenoplectus lacustris*.

@@

18. *Convolvulo arvensis-Agropyrion repentis* Semiruderal grasslands. Occupies wetter and shadier situations than the other alliances in the class, and has a higher proportion of geophytes and grasses (**Fig. 4**). Isolated occurrence in the south of the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Poa compressa*, *Elytrigia repens*, *Potentilla reptans*, *Valerianella locusta*.

19. *Carduo carpetani-Cirsion odontolepidis* Thistle ruderal vegetation of warm sites. Occupies warmer situations than the other thistle alliance in the class (**Fig. 4**). Mostly distributed in the Cantabrian Mountains (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Onopordum acanthium*.

20. *Cirsion richterano-chodati* Thistle ruderal vegetation of cold sites. Occupies colder situations than the other thistle alliance in the class (**Fig. 4**). Mostly distributed in the Cantabrian Mountains (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Cirsium eriophorum*, *Carduus carpetanus*, *Carduus nutans* aggr.

21. *Dauco-Melilotion* Biennial ruderal vegetation. Occupies wetter situations than the two thistle alliances in the class, and has a higher proportion of therophytes (**Fig. 4**). Sparse occurrence throughout the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Helminthotheca echioides*, *Daucus carota*, *Melilotus albus*, *Foeniculum vulgare*.

@@

22. *Geo urbani-Alliarion officinalis* Low-herb short-lived semiruderal and fringe vegetation. Occupies shadier and less disturbed situations than the other alliances in the class, and has a higher proportion of therophytes and shorter plants (**Fig. 4**). Wide distribution in the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Urtica dioica*.

23. *Arction lappae* Low-herb short-lived ruderal vegetation. Occupies sunnier and more frequently-disturbed situations than the other alliances in the class, and has a higher proportion of therophytes and shorter plants (**Fig. 4**). Distribution concentrated in the Cantabrian Mountains (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Oxybasis rubra*, *Senecio duriaei*.

24. *Balloto-Conion maculati* Tall-herb perennial ruderal vegetation. Occupies warmer situations than the other alliances in the class (**Fig. 4**). Distribution concentrated in the Cantabrian Mountains and the east of the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Sambucus ebulus*.

25. *Aegopodion podagrariae* Tall-herb perennial vegetation of forest margins and clearings. Occupies shadier and less disturbed situations than the other alliances in the class (**Fig. 4**). Sparse occurrence throughout the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Anthriscus sylvestris*, *Heracleum sphondylium*, *Galium aparine*.

26. *Epilobion angustifolii* Tall-herb perennial vegetation of forest margins and clearings in acidic soils. Occupies colder, nutrient-poorer, more acidic and less disturbed situations than the other alliances in the class (**Fig. 4**). Distributed towards the west of the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Epilobium angustifolium*, *Digitalis purpurea*, *Luzula lactea*.

27. *Cynancho-Convolvulion sepium* Tall-herb vegetation of nutrient-rich riparian habitats. Occupies wetter situations and less disturbed situations than the other alliances in the class, and has a later flowering season (**Fig. 4**). Dispersed throughout the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Eupatorium cannabinum*, *Angelica sylvestris*, *Picris hieracioides*.

28. *Senecionion fluviatilis* Tall-herb vegetation of nutrient-rich river banks and ditches. Occupies warmer and wetter situations than the other alliances in the class, flowers later, has taller plants and is highly dominated by neophytes and geophytes (**Fig. 4**). Isolated occurrence in the western coast of the ecoregion (**Fig. 1**). Diagnostic species (**Supplementary 5**): *Arundo donax*, *Silene latifolia*.

@@

Compare proportion of neophytes to other areas [@RN5429; \@ @RN5435]

@@

History of synanthropic veg in czech [@RN5496]

@@

The original datasets, as well as R code for analysis and creation of the manuscript can be accessed at the GitHub repository <https://github.com/efernandezpascual/manmade>. Upon publication, a version of record of the repository will be deposited in Zenodo.

@@

```{r fig1, echo = FALSE, fig.pos = "H", fig.cap = "Figure 1. Anthropogenic vegetation alliances of the Iberian Atlantic ecoregion. Each dot is a vegetation plot. Dot colors indicate the vegetation class."}
knitr::include_graphics("../results/figures/F1 - maps.png")
```

```{r fig2, echo = FALSE, fig.pos = "H", fig.cap = "Figure 2. Floristic composition of the anthropogenic plant communities of the Iberian Atlantic ecoregion. Biplots produced by Principal Component Analysis (PCA) with Hellinger transformation. PCAs were conducted for the whole vegetation (A) and for specific subsets (B-F). Colors indicate the vegetation class."}
knitr::include_graphics("../results/figures/F2 - pca.png")
```

```{r fig3, echo = FALSE, fig.pos = "H", fig.cap = "Figure 3. Proportion of neophytes, archaeophytes and native species in the anthropogenic plant communities of the Iberian Atlantic ecoregion. Each bar represents the plot-level average proportion of neophytes (dark shade), archaeophytes (medium shade) and natives (light shade) in each vegetation alliance. Colors indicate the vegetation class."}
knitr::include_graphics("../results/figures/F3 - origins.png")
```

```{r fig4, echo = FALSE, fig.pos = "H", fig.cap = "Figure 4. Proportion of therophytes, geophytes and other life forms in the anthropogenic plant communities of the Iberian Atlantic ecoregion. Each bar represents the plot-level average proportion of therophytes (dark shade), geophytes (medium shade) and other life forms (light shade) in each vegetation alliance. Colors indicate the vegetation class."}
knitr::include_graphics("../results/figures/F4 - lifeforms.png")
```

```{r fig5, echo = FALSE, fig.pos = "H", fig.cap = "Figure 5. Vegetation heigh and flowering phenology of the anthropogenic plant communities of the Iberian Atlantic ecoregion. Boxplots show the community-weighted mean for plant heigh, month of peak flowering and length of the flowering season for each vegetation alliance. Colors indicate the vegetation class."}
knitr::include_graphics("../results/figures/F5 - height-flowering.png")
```

```{r fig6, echo = FALSE, fig.pos = "H", fig.cap = "Figure 6. Ecological requirements of the anthropogenic plant communities of the Iberian Atlantic ecoregion. Boxplots show the plot-level means for the ecological indicator values of temperature, moisture, light, nutrients, soil reaction, disturbance frequency and disturbance severity, calculated for each vegetation alliance. Colors indicate the vegetation class."}
knitr::include_graphics("../results/figures/F6 - indicators.png")
```

```{r fig7, echo = FALSE, fig.pos = "H", fig.cap = "Figure 7. Pattern of variation in the ecological requirements of the anthropogenic plant communities of the Iberian Atlantic ecoregion. Biplot produced by Principal Component Analysis (PCA) of the plot-level means for the ecological indicator values of temperature, moisture, light, nutrients, soil reaction, disturbance frequency and disturbance severity. Labels and arrows indicate the contribution of each indicator to the first and second principal components. Colors indicate the vegetation class."}
knitr::include_graphics("../results/figures/F7 - pca.png")
```
